
\section{Quotient Classes}
\label{sec:quotients}

Previously, we demonstrated how to group and aggregate data---so long as the
structure of the data reflects the hierarchical form of the database.  In this
section, we show how to overcome this limitation.

In Figure~\ref{fig:unfolded-form}, the schema graph is unfolded into an
infinite tree, shaping the data into a hierarchical form.  A section of this
hierarchy could be extracted using the $\Select$ combinator.

\begin{demo}
    \label{ex:department-select-name-employee}
    Show all departments, and, for each department, list the associated
    employees.
    \begin{equation*}
        \Department\Apply\Select(\Name,\; \Employee)
    \end{equation*}
\end{demo}
But what if we ask for \emph{positions} instead of \emph{departments}?

\begin{demo}
    \label{ex:employee-group-position}
    Show all positions, and, for each position, list the associated employees.
\end{demo}

Unlike the previous example, this query does not match the structure of the
database and, therefore, cannot be constructed as easily.  Indeed,
Example~\ref{ex:department-select-name-employee} is built from the primitives
\begin{alignat*}{3}
    & \Department && : \Void && \to \Seq{\Dept}, \\
    & \Name && : \Dept && \to \Text, \\
    & \Employee && : \Dept && \to \Seq{\Emp}.
\end{alignat*}
To construct Example~\ref{ex:employee-group-position} in a similar fashion, we
need a hypothetical class $\Pos$ of \emph{position} entities and a set of
queries with the corresponding signatures
\begin{equation}
    \label{eq:group-components}
    \begin{alignedat}{2}
        & \Void && \to \Seq{\Pos}, \\
        & \Pos && \to \Text, \\
        & \Pos && \to \Seq{\Emp}.
    \end{alignedat} \tag{$\star\star\star$}
\end{equation}
However, there is no built-in class of position entities and we only have the
following primitives available:
\begin{alignat*}{3}
    & \Employee && : \Void && \to \Seq{\Emp}, \\
    & \Position && : \Emp && \to \Text.
\end{alignat*}

To make a ``virtual'' entity class from all distinct values of an attribute and
inject this class into the database structure, we use the $\Group$ combinator.
For example (see Figure~\ref{fig:group-combinator}), a list of \emph{all
distinct employee positions} can be produced with the query
\begin{equation*}
    \Employee\Apply\Group(\Position) : \Void \to \Seq{\Pos}.
\end{equation*}
The virtual $\Pos$ class comes with the primitives
\begin{alignat*}{3}
    & \Position && : \Pos && \to \Text, \\
    & \Employee && : \Pos && \to \Seq{\Emp},
\end{alignat*}
which, \emph{given a position entity, produce respectively the position name
and a list of associated employees}.
This gives us all the query components (see~(\ref{eq:group-components}) above)
needed to complete the example.

\input{figure-group-combinator}

\addtocounter{demo}{-1}
\begin{demo}
    Show all positions, and, for each position, list the associated employees.
    \begin{align*}
        & \Employee \\
        & \Apply\Group(\Position) \\
        & \Apply\Select(\Position,\; \Employee)
    \end{align*}
\end{demo}

The query
\begin{equation*}
    \Employee\Apply\Group(\Position)
\end{equation*}
correlates all distinct values emitted by $\Position$ with the respective
$\Employee$ entities and packs them together into the records of type
\begin{equation*}
    \Pos \equiv \Tuple{\Position : \Text,\; \Employee : \Seq{\Emp}}.
\end{equation*}
We call $\Pos$ a \emph{quotient class} and denote it by
\begin{equation*}
    \Quotient{\Emp}{\Position}.
\end{equation*}

Once the database hierarchy is rearranged to include the class $\Pos$, we can
answer any questions about position entities.

\begin{demo}
    \label{ex:employee-group-position-etc}
    In the Police department, show all positions with the number of employees
    and the top salary.
    \begin{alignat*}{2}
        & \Employee\hidewidth && \\
        & \Apply\Filter(\Department\To\Name = \textliteral{POLICE})\hspace{-15em}&&\hspace{12em} \\
        & \Apply\Group(\Position)\hidewidth && \\
        & \Apply\Select(&& \Position, \\
        & && \Count(\Employee), \\
        & && \Max(\Employee\To\Salary))
    \end{alignat*}
\end{demo}

Here, for each position in the Police department, we determine two calculated
attributes, the number of employees and the top salary:
\begin{alignat*}{3}
    & \Count(\Employee) && : \Quotient{\Emp}{\Position} && \to \Int, \\
    & \Max(\Employee\To\Salary) && : \Quotient{\Emp}{\Position} && \to \Opt{\Int}.
\end{alignat*}

\begin{demo}
    \label{ex:nested-group}
    Arrange employees into a hierarchy: first by position, then by department.
    \begin{alignat*}{2}
        & \Employee\hidewidth && \\
        & \Apply\Group(\Position)\hidewidth && \\
        & \Apply\Select(&& \Position, \\
        & && \Employee \\
        & && \Apply\Group(\Department) \\
        & && \Apply\Select(\Department\To\Name,\; \Employee))
    \end{alignat*}
\end{demo}

Nested $\Group$ combinators can construct a hierarchical output of an arbitrary
form.  In this example, we rebuild the database hierarchy to place positions on
top, then departments, and then employees.  Notably, the nested $\Group$
expression has a signature
\begin{multline*}
    \Employee\Apply\Group(\Department) : \\
    \Quotient{\Emp}{\Position} \to \Seq{\Quotient{\Emp}{\Department}}.
\end{multline*}

\begin{demo}
    \label{ex:unique-department}
    Show all positions available in more than one department, and, for each
    position, list the respective departments.
    \begin{alignat*}{2}
        & \Employee\hidewidth &&  \\
        & \Apply\Group(\Position)\hidewidth && \\
        & \Apply\Define(&& \Department\As \\
        & && \qquad \Unique(\Employee\To\Department)) \\
        & \Apply\Filter(\Count(\Department) > 1)\hidewidth && \\
        & \Apply\Select(\Position,\; \Department\To\Name)\hidewidth &&
    \end{alignat*}
\end{demo}

This example uses the $\Unique$ combinator to find all distinct entities in a
list of departments.  The $\Unique$ combinator can be expressed via $\Group$ by
forgetting the plural component of the quotient class.  In this example,
$\Unique(\Employee\To\allowbreak\Department)$ is equivalent to
\begin{equation*}
    \Employee\Apply\Group(\Department)\To\Department.
\end{equation*}

\begin{demo}
    \label{ex:employee-group-level}
    How many employees at each level of the organization chart?
    \begin{align*}
        & \Employee \\
        & \Apply\Group(\Level\As\Count(\Connect(\Manager))) \\
        & \Apply\Select(\Level,\; \Count(\Employee))
    \end{align*}
\end{demo}

In order to apply $\Group$ to a calculated attribute, such as \emph{the level
in the organization chart}
\begin{equation*}
    \Count(\Connect(\Manager)) : \Emp \to \Int,
\end{equation*}
we need to bind an explicit tag to this attribute.

\begin{demo}
    \label{ex:rollup}
    Show the average salary by department and position, with subtotals for each
    department and the grand total.
    \begin{alignat*}{2}
        & \Employee\hidewidth && \\
        & \Apply\Rollup(\Department,\; \Position)\hidewidth && \\
        & \Apply\Select(&& \Department, \\
        & && \Position, \\
        & && \Mean(\Employee\To\Salary))
    \end{alignat*}
\end{demo}

To summarize data along several dimensions, we can apply $\Group$ to more than
one attribute.  When the summary data has to include subtotals and totals, we
replace $\Group$ with $\Rollup$.

In this example, the query
\begin{equation*}
    \Employee\Apply\Rollup(\Department,\; \Position)
\end{equation*}
produces a sequence of records of type
\begin{multline*}
    \Quotient{\Emp}{\Department_{\bot},\Position_{\bot}} \equiv \\
    \begin{alignedat}{3}
        & \Tuple{&& \Department && : \Opt{\Dept}, \\
        & && \Position && : \Opt{\Text}, \\
        & && \Employee && : \Seq{\Emp}}.
    \end{alignedat}
\end{multline*}
In addition to the records that would be generated by $\Group$, $\Rollup$ emits
one ``subtotal'' record per each department and one ``grand total'' record.
The former has the $\Position$ field set to $\bot$ and an $\Employee$ list
containing all employees in the given department.  The latter has both
$\Department$ and $\Position$ set to $\bot$ and $\Employee$ containing the full
list of employees.

